---
title: "Data wrangling"
author: "Victor Kibisu"
date: '2022-05-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

So in this lesson we will focus on data wrangling using dplyr, therefore the folloowing are the objectives of this training:
Learning Objectives.

0. What is dplyr package?
1.Select columns in a data frame with the dplyr function select.
2.Select rows in a data frame according to filtering conditions with the dplyr function filter.
3.Direct the output of one dplyr function to the input of another function with the ‘pipe’ operator %>%.
4.Add new columns to a data frame that are functions of existing columns with mutate.
5.Understand the split-apply-combine concept for data analysis.
6.Use summarize, group_by, and count to split a data frame into groups of observations, apply a summary statistics for each group, and then combine the results.
7.Join two tables by a common variable.

### download the data
```{r}
### Download MS_trafficstops_bw_age.csv
download.file("http://bit.ly/MS_trafficstops_bw_age",
              "MS_trafficstops_bw_age.csv")

## Download MS_acs2015_bw.csv:
download.file("http://bit.ly/MS_acs_2015_bw",
              "MS_acs2015_bw.csv")


```

### 1.1 What is dplyr?
dplyr is one part of a larger tidyverse that enables you to work with data in tidy data formats. “Tidy datasets are easy to manipulate, model and visualise, and have a specific structure: each variable is a column, each observation is a row, and each type of observational unit is a table.

### 1.2 Subsetting columns and rows
We will be working on a small subset of the data from the Stanford Open Policing Project. It contains information about traffic stops for blacks and whites in the state of Mississippi during January 2013 to mid-July of 2016.

### Import data
```{r}
stops <- read_csv("MS_trafficstops_bw_age.csv")
```

```{r}
## check for the first few observations
head(stops)
```

```{r}
## structure of the data
str(stops)
```
#### Select variables
```{r}
sub1 <-stops %>%  select(police_department, officer_id, driver_race)


## use of helper functions
driver_xt <- select(stops, starts_with("driver"))
```


To subset rows based on specific criteria, we use filter():
```{r}
yazoo <- stops %>% filter(county_name == "Yazoo")
```

Here are some other ways to subset rows:
```{r}
#by row number: 
slice(stops, 1:3) # rows 1-3
```

```{r}
#rows with highest or lowest values of a variable
slice_min(stops, driver_age) 
slice_max(stops, driver_age)# likewise slice_max()


```

```{r}
## random rows
slice_sample(stops, n = 5) # number of rows to select
slice_sample(stops, prop = .0001) # fraction of rows to select
```

We might also be intrested in sorting your data based on a variable
```{r}
arrange(stops, county_name, stop_date)
```

### 1.3 Pipes
What if you wanted to filter and select on the same data? For example, lets find drivers over 85 years and only keep the violation and gender columns. There are several methods that can be used to do this one is by intermediate steps:
```{r}
tmp_df <- filter(stops, driver_age > 85)
select(tmp_df, violation, driver_gender)
```

Nested functions
```{r}
select(filter(stops, driver_age > 85), violation, driver_gender)
```

And pipes
```{r}
stops %>%
  filter(driver_age > 85) %>%
  select(violation, driver_gender)
```

```{r}
senior_drivers <- stops %>%
  filter(driver_age > 85) %>%
  select(violation, driver_gender, driver_race)

senior_drivers
```

### challenge
Using pipes, subset the stops data to include stops in Tunica county only and retain the columns stop_date, driver_age, and violation. Bonus: sort the table by driver age.

### 1.4 Add new columns
Frequently you’ll want to create new columns based on the values in existing columns or. For this we’ll use mutate(). We can also reassign values to an existing column with that function.
```{r}
stops %>% 
  mutate(birth_year = year(driver_birthdate))

stops %>% 
  mutate(birth_year = year(driver_birthdate),
         birth_cohort = floor(birth_year/10)*10) 
```

the power of pipe operator
```{r}
stops %>% 
  mutate(birth_year = year(driver_birthdate),
         birth_cohort = floor(birth_year/10)*10,
         birth_cohort = factor(birth_cohort)) %>%
    select(birth_cohort) %>% 
    plot()
```


Use of ifelse
```{r}
stops %>% 
  mutate(cohort = ifelse(year(driver_birthdate) < 2000, "pre-millenial", "millenial")) %>% 
  select(driver_birthdate, cohort)
```


